---
title: "Pf6 metadata"
author: "Aris Paschalidis"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include = FALSE}
here::i_am("metadata/pf6_meta.Rmd")

library(ggplot2)
```

This file outlines how we compile the metadata for the Pf6 dataset. We first
download the most basic metadata from the Pf6 website.

```{r initial meta}
temp_file <- tempfile()
download.file(
  "ftp://ngs.sanger.ac.uk/production/malaria/pfcommunityproject/Pf6/Pf_6_samples.txt",
  temp_file
)
meta <- data.table::fread(temp_file)
```

We can then cluster our samples into suitably close regions, for which we assume
the same PLMAF exists but without reducing our $n$. We can use a silhouette plot
to visualize the optimum number of regions to guide our decisions.

```{r k-means}
# K Means
locations <- unique(meta[, c("Lat", "Long")])

sis <- 2:(nrow(locations) - 1)
for (k in sis) {
  si <- cluster::silhouette(cluster::pam(x = locations, k))
  sis[k - 1] <- mean(si[, 3])
}

# Silhouette plot
# N.B. element_line throws a warning when give it a list as an input. A
# workaround for this for making the number a different color is to use
# `ggtext::element_markdown`, but this does not work for making the tick mark
# another color
sil <- data.frame(
  cluster = seq_along(sis) + 1,
  sis = sis
)
ggplot(sil, aes(x = cluster, y = sis)) +
  geom_point() +
  geom_vline(xintercept = which(diff(sign(diff(sis))) == -2) + 2, linetype = 5) +
  geom_vline(xintercept = 24, linetype = 5, color = "red") +
  annotation_custom(
    grob = grid::segmentsGrob(gp = grid::gpar(col = "red", lwd = 2)),
    xmin = 24, xmax = 24, ymin = -0.025, ymax = -0.02
  ) +
  scale_x_continuous(breaks = c(0, 20, 24, 40, 60)) +
  labs(x = "Number of Clusters", y = "Average Silhoutte Score") +
  coiaf::theme_coiaf() +
  theme(
    axis.title = element_text(size = 8),
    axis.text.x = ggtext::element_markdown(color = c("black", "black", "red", "black", "black")),
    axis.ticks.x = element_line(color = c("black", "black", "red", "black", "black"))
  )
```

Based on this plot, we select 24 regions as the optimal cluster number. We then
cluster our samples and combine the data with our predictions.

```{r cluster}
# Clustering with 24 locations
ks <- cluster::pam(meta[, c("Lat", "Long")], k = 24)
meta$color <- as.factor(ks$clustering)
```

We may also be interested in the prevalence of malaria in various regions and
the FwS metric.

```{r prevalence data, message = FALSE, warning = FALSE}
# Get the lat longs
coords <- data.frame(meta$Long, meta$Lat)
names(coords) <- c("x", "y")

# make column for 2-10 year old prevalence
meta$prev_2_10 <- as.numeric(meta$Year)
meta <- tidyr::drop_na(meta)

# get rasters using malariaAtlas package (this will take like 15 mins to download)
PfPR2_10 <- malariaAtlas::getRaster(year = sort(unique(meta$prev_2_10)))

# loop through and extra malaria prevalence
for (i in seq_along(sort(unique(meta$Year)))) {
  year <- sort(unique(meta$Year))[i]
  pos <- which(meta$Year == year)
  i_coords <- coords[pos, ]
  prev <- raster::extract(PfPR2_10[[i]], i_coords)
  meta$prev_2_10[pos] <- prev
}

meta
```

```{r fws data}
fws <- read.csv("ftp://ngs.sanger.ac.uk/production/malaria/pfcommunityproject/Pf6/Pf_6_fws.txt", sep = "\t")
fws
```

Finally, we can combine all our metadata together and save.

```{r final metadata}
combine_meta <- dplyr::left_join(meta, fws, by = "Sample")

saveRDS(combine_meta, here::here("metadata", "pf6_meta.rds"))
```
