---
title: "Pf6 metadata"
author: "Aris Paschalidis"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include = FALSE}
here::i_am("metadata/pf6_meta.Rmd")

library(ggplot2)
```

This file outlines how we compile the metadata for the Pf6 dataset. We first
download the most basic metadata from the Pf6 website.

```{r initial meta}
temp_file <- tempfile()
download.file(
  "ftp://ngs.sanger.ac.uk/production/malaria/pfcommunityproject/Pf6/Pf_6_samples.txt",
  temp_file
)
meta <- data.table::fread(temp_file)
```

One of the key pieces of information we will later utilize is the country in
which each sample was obtained. In some cases, it may be more useful to look
on a broader scale at continents or subregions of sample.
```{r country region}
meta <- dplyr::mutate(
  meta,
  Continent = countrycode::countrycode(
    Country, 
    origin = "country.name",
    destination = "continent"
  ),
  Subregion = countrycode::countrycode(
    Country, 
    origin = "country.name",
    destination = "region23"
  ),
  .before = Country
)
```

We can then cluster our samples into suitably close regions, for which we assume
the same PLMAF exists but without reducing our $n$. We can use a silhouette plot
to visualize the optimum number of regions to guide our decisions.

```{r k-means}
# K Means
locations <- unique(meta[, c("Lat", "Long")])

sis <- 2:(nrow(locations) - 1)
for (k in sis) {
  si <- cluster::silhouette(cluster::pam(x = locations, k))
  sis[k - 1] <- mean(si[, 3])
}

# Silhouette plot
# N.B. element_line throws a warning when give it a list as an input. A
# workaround for this for making the number a different color is to use
# `ggtext::element_markdown`, but this does not work for making the tick mark
# another color
sil <- data.frame(
  cluster = seq_along(sis) + 1,
  sis = sis
)
ggplot(sil, aes(x = cluster, y = sis)) +
  geom_point() +
  geom_vline(xintercept = which(diff(sign(diff(sis))) == -2) + 2, linetype = 5) +
  geom_vline(xintercept = 24, linetype = 5, color = "red") +
  annotation_custom(
    grob = grid::segmentsGrob(gp = grid::gpar(col = "red", lwd = 2)),
    xmin = 24, xmax = 24, ymin = -0.025, ymax = -0.02
  ) +
  scale_x_continuous(breaks = c(0, 20, 24, 40, 60)) +
  labs(x = "Number of Clusters", y = "Average Silhoutte Score") +
  coiaf::theme_coiaf() +
  theme(
    axis.title = element_text(size = 8),
    axis.text.x = ggtext::element_markdown(color = c("black", "black", "red", "black", "black")),
    axis.ticks.x = element_line(color = c("black", "black", "red", "black", "black"))
  )

# Save figure
ggsave(
  here::here("figures", "silhoutte.png"),
  device = "png", 
  width = 2250, 
  height = 1500,
  units = "px", 
  dpi = "print"
)
```

Based on this plot, we select 24 regions as the optimal cluster number. We then
cluster our samples and combine the data with our predictions.

```{r cluster}
# Clustering with 24 locations
ks <- cluster::pam(meta[, c("Lat", "Long")], k = 24)
meta$color <- as.factor(ks$clustering)
```

```{r plot cluster}
ggplot(meta) +
  borders("world") +
  geom_polygon(
    data = map_data("world"),
    aes(x = long, y = lat, group = group),
    fill = "grey",
    alpha = 0.3
  ) +
  geom_point(aes(x = Long, y = Lat, color = color), alpha = 0.8) +
  theme_void() +
  theme(legend.position = "none") +
  scale_size(guide = "none") +
  coord_quickmap(xlim = c(-75, 150), ylim = c(-30, 30))

# Save figure
ggsave(
  here::here("figures", "cluster-locations.png"),
  device = "png", 
  width = 2250, 
  height = 600,
  units = "px", 
  dpi = "print"
)
```

We may also be interested in the prevalence of malaria in various regions and
the FwS metric. We extract the prevalence data from malaria atlas.

```{r prevalence data, message = FALSE, warning = FALSE}
# Get the lat longs
coords <- data.frame(meta$Long, meta$Lat)
names(coords) <- c("x", "y")

# make column for 2-10 year old prevalence
meta$prev_2_10 <- as.numeric(meta$Year)
meta <- tidyr::drop_na(meta)

# get rasters using malariaAtlas package (this will take like 15 mins to download)
PfPR2_10 <- malariaAtlas::getRaster(
  surface = "Plasmodium falciparum PR2 - 10 version 2020", 
  year = sort(unique(meta$prev_2_10))
)

# loop through and extra malaria prevalence
for (i in seq_along(sort(unique(meta$Year)))) {
  year <- sort(unique(meta$Year))[i]
  pos <- which(meta$Year == year)
  i_coords <- coords[pos, ]
  prev <- raster::extract(PfPR2_10[[i]], i_coords)
  meta$prev_2_10[pos] <- prev
}

meta
```

And extract the FwS data from the Pf6 datasets.

```{r fws data}
fws <- read.csv("ftp://ngs.sanger.ac.uk/production/malaria/pfcommunityproject/Pf6/Pf_6_fws.txt", sep = "\t")
fws
```

Additionally, we may be interested in the effective K metric defined in the
manuscript accompanying the DEploidIBD software. We download the data from 
[GitHub](https://github.com/mcveanlab/mixedIBD-Supplement/blob/master/compiled_data_corrected.Rdata)
and read it in.

```{r effective K}
deploid <- get(load(here::here("metadata", "deploidibd_data.rdata")))
deploid_eff_k <- dplyr::select(deploid, ID, eff_k)
```

Finally, we can combine all our metadata together and save.

```{r final metadata}
combine_meta <- meta %>% 
  dplyr::left_join(fws, by = "Sample") %>% 
  dplyr::left_join(deploid_eff_k, by = c("Sample" = "ID"))

saveRDS(combine_meta, here::here("metadata", "pf6_meta.rds"))
```
