---
output: github_document
bibliography: references/references.bib
csl: references/plos-computational-biology.csl
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```

# Real Data Analysis With coiaf

<!-- badges: start -->
[![Requirement](https://img.shields.io/badge/requirement-coiaf-blue)](https://github.com/bailey-lab/coiaf)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

This repository stores the real data analysis conducted to test the software
package [{coiaf}](https://github.com/bailey-lab/coiaf).

## Data Source
We analysed samples from the MalariaGEN *Plasmodium falciparum* Community
Project [@malariagen_open_2021]. The MalariaGEN *Plasmodium falciparum*
Community Project provides genomic data from over 7,000 *P. falciparum* samples
from 28 malaria-endemic countries in Africa, Asia, South America, and Oceania
from 2002-2015. Detailed information about the data release including brief
descriptions of contributing partner studies and study locations is available in
the supplementary of MalariaGEN *et al.*.

## Project Structure

```{r proj structure, echo = FALSE}
fs::dir_tree(
  regexp = "renv.*|library|.*figures|archive|.*Rproj|.*E.R?md|.*html|data-outputs/vary-seq-error/*|references", 
  invert = TRUE
)
```

## References
