---
title: "Comparison of estimation strategies"
author: "Aris Paschalidis"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include = FALSE}
# Here, we set default options for our markdown file
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(coiaf)
library(ggplot2)
library(patchwork)
```

```{r base pf6}
base_pf6 <- readRDS(here::here("data-outputs", "base_pf6.rds"))
```

```{r freq variation}
freq_variation <- readRDS(here::here("data-outputs", "freq_variation.rds")) %>% 
  dplyr::filter(
    !is.na(dis_var_med), 
    !is.na(dis_freq_med), 
    !is.na(cont_var_med), 
    !is.na(cont_freq_med), 
    !is.na(Region)
  )
```

```{r intersecting wsafs}
intersecting_wsafs <- readRDS(here::here(
  "data-outputs", 
  "intersecting_wsafs.rds"
))

rmcl_predictions <- readRDS(here::here(
  "data-outputs",
  "rmcl_estimation.rds"
))

intersecting_wsafs <- dplyr::left_join(
  intersecting_wsafs, 
  rmcl_predictions, 
  by = c("name", "Region" = "region")
)
```

```{r intersecting wsafs low error}
regional_wsafs <- readRDS(here::here(
  "data-outputs", 
  "regional_wsafs.rds"
))

regional_wsafs <- dplyr::left_join(
  regional_wsafs, 
  rmcl_predictions, 
  by = c("name", "Region" = "region")
)
```

We can see that there is a difference in the number of samples between our data
sets. A more in depth investigation yields that there are 4 samples that are not
in the `freq_variation` results: "PW0003-C", "PW0004-C", "PW0014-C", and
"PW0016-C".

```{r differing region}
freq_count <- dplyr::count(freq_variation, Region)

inter_count <- intersecting_wsafs %>% 
  dplyr::count(Region) %>% 
  dplyr::arrange(as.numeric(Region))

which(freq_count$n != inter_count$n)
```

```{r differing samples}
freq_name <- freq_variation %>% 
  dplyr::filter(Region == 19) %>% 
  dplyr::pull(name)

inter_name <- intersecting_wsafs %>% 
  dplyr::filter(Region == 19) %>% 
  dplyr::pull(name)

inter_name[(!inter_name %in% freq_name)]
```

## Understanding relationships

```{r combine all the estimations}
inter_bind <- intersecting_wsafs %>% 
  dplyr::mutate(Region = as.numeric(Region)) %>% 
  dplyr::rename(
    dis_var_med = dis_var, 
    dis_freq_med = dis_freq, 
    cont_var_med = cont_var, 
    cont_freq_med = cont_freq
  )

regional_bind <- regional_wsafs %>% 
  dplyr::mutate(Region = as.numeric(Region)) %>% 
  dplyr::rename(
    dis_var_med = dis_var, 
    dis_freq_med = dis_freq, 
    cont_var_med = cont_var, 
    cont_freq_med = cont_freq
  )

combined_data <- dplyr::bind_rows(
  list(base = base_pf6, variation = freq_variation, inter = inter_bind, regional = regional_bind), 
  .id = "dataset"
)

no_base_combined <- dplyr::bind_rows(
  list(variation = freq_variation, inter = inter_bind, regional = regional_bind), 
  .id = "dataset"
)
```

Between estimation techniques.

```{r relationship between techqiues}
plot_bar <- function(data, col, xlab) {
 ggplot(data, aes({{ col }}, fill = dataset)) +
  geom_bar(position = "dodge") +
  scale_fill_viridis_d() +
  labs(x = xlab, y = "Count", fill = "Dataset") +
  theme_coiaf()
}

plot_density <- function(data, col, xlab) {
 ggplot(data, aes({{ col }}, fill = dataset)) +
  geom_density() +
  facet_wrap(~dataset) +
  scale_fill_viridis_d() +
  labs(x = xlab, y = "Density", fill = "Dataset") +
  theme_coiaf()
}

method_comparison <- function(data) {
  pdv <- plot_bar(data, dis_var_med, "Discrete Variant Method")
  pcv <- plot_density(data, cont_var_med, "Continuous Variant Method")

  pdf <- plot_bar(data, dis_freq_med, "Discrete Frequency Method")
  pcf <- plot_density(data, cont_freq_med, "Continuous Frequency Method")
  
  pdv + pcv + pdf + pcf + plot_layout(guides = "collect")
}

rmcl_helper <- function(data, col) {
  title <- switch(col,
    dis_var_med = "Discrete Variant Method",
    dis_freq_med = "Discrete Frequency Method",
    cont_var_med = "Continuous Variant Method",
    cont_freq_med = "Continuous Frequency Method"
  )
  
  ggplot(data, aes(x = rmcl_med, y = .data[[col]], color = dataset)) +
    facet_wrap(~dataset) +
    geom_abline(color = "gray", size = 0.5) +
    geom_jitter(alpha = 0.5) +
    theme_bw() +
    theme(
      panel.grid.minor = element_blank(), 
      plot.title = element_text(hjust = 0.5, size = 12), 
      legend.title = element_text(size = 12), 
      legend.position = "right"
    ) +
    guides(size = "none") +
    scale_x_continuous("THE REAL McCOIL Prediction", breaks = seq(0, 10)) +
    labs(title = title, y = "coiaf Prediction", color = "Dataset")
}

rmcl_comparison <- function(data) {
  data = dplyr::filter(data, rmcl_med != 25)
  
  rmcl_helper(data, "dis_var_med") + 
    rmcl_helper(data, "dis_freq_med") +
    rmcl_helper(data, "cont_var_med") +
    rmcl_helper(data, "cont_freq_med") + 
    plot_layout(guides = "collect")
}
```
```

Compared to THE REAL McCOIL

```{r rmcl examination}
ggplot(
  data = dplyr::filter(combined_data, rmcl_med != 25), 
  mapping = aes(x = rmcl_med, y = dis_var_med, color = dataset)
) +
  facet_wrap(~dataset) +
  geom_abline(color = "gray", size = 0.5) +
  geom_jitter(alpha = 0.5) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  guides(size = "none") +
  scale_x_continuous("THE REAL McCOIL Prediction", breaks = seq(0, 10)) +
  labs(title = "Variant Method", y = "coiaf Prediction")

ggplot(
  data = dplyr::filter(combined_data, rmcl_med != 25), 
  mapping = aes(x = rmcl_med, y = dis_freq_med, color = dataset)
) +
  facet_wrap(~dataset) +
  geom_abline(color = "gray", size = 0.5) +
  geom_jitter(alpha = 0.5) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  guides(size = "none") +
  scale_x_continuous("THE REAL McCOIL Prediction", breaks = seq(0, 10)) +
  labs(title = "Frequency Method", y = "coiaf Prediction")
```

## Isolating samples where we deviate from RMCL

First find samples where RMCL predicts a high COI.

```{r high coi samples}
high_samples <- combined_data %>% 
  dplyr::filter(dataset == "inter", rmcl_med > 5, rmcl_med != 25) %>% 
  dplyr::select(name, Region)
```

Plot the sample

```{r plot the sample}
path <- "~/Desktop/Malaria/COI data/"
rmcl_wsafs <- readRDS(paste0(path, "RMCL_wsafs_unique.rds"))

purrr::walk(seq_len(nrow(high_samples)), function(i) {
  if (i %% 9 == 1) {
    pdf(paste0("base_samples_", i, ".pdf"))
    par(mfrow = c(3, 3))
  }
  
  sample_name <- high_samples$name[i]
  sample_region <- high_samples$Region[i]
  
  # path <- "~/Desktop/Malaria/COI data/new-wsafs/intersecting-regions/"
  # wsaf_matrix <- readRDS(paste0(path, glue::glue("region_{sample_region}.rds")))
  
  # Get region matrix
  wsaf_matrix <- names(rmcl_wsafs) %>%
    stringr::str_detect(stringr::str_c("cat_region_", sample_region, "_vcf_0")) %>%
    purrr::keep(rmcl_wsafs, .)
  wsaf_matrix <- wsaf_matrix[[1]]
  
  plaf <- colMeans(wsaf_matrix, na.rm = T)
  wsaf <- wsaf_matrix[sample_name, ]
  input <- tibble::tibble(wsaf = wsaf, plaf = plaf) %>% tidyr::drop_na()
  input$plaf[input$plaf > 0.5] <- 1 - input$plaf[input$plaf > 0.5]
  
  # het_data <- dplyr::filter(input, wsaf > 0 & wsaf < 1)
  plot(input$plaf, input$wsaf, main = sample_name)
  
  # Compute windowed averages
  df_grouped <- process_real(
    input$wsaf, input$plaf, 
    bin_size = 50, 
    coi_method = "frequency"
  )

  # What was the seq error?
  df_grouped$seq_error
  
  # Add average to plot
  lines(df_grouped$data$midpoints, df_grouped$data$m_variant,
    col = "blue", lwd = 2, type = "o", pch = 20
  )
  
  
  if (i %% 9 == 0 | i == nrow(high_samples)) dev.off()
})

# dplyr::filter(combined_data, name == sample_name, Region == sample_region)
```


