---
title: "Comparison of estimation strategies"
author: "Aris Paschalidis"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include = FALSE}
# Here, we set default options for our markdown file
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(coiaf)
library(ggplot2)
library(patchwork)
library(magrittr)
```

This analysis file will be used to compare the different estimation strategies
we employ to examine the COI in the world. The COI estimations we compare have
all been generated from the filtering process illustrated in the [download data
folder](https://github.com/bailey-lab/coiaf-real-data/tree/main/download).

## Loading estimations

```{r rmcl estimation}
rmcl_predictions <- readRDS(here::here(
  "data-outputs",
  "rmcl_estimation.rds"
))
```

```{r core genome}
core <- readRDS(here::here(
  "data-outputs", 
  "core-genome.rds"
)) %>% 
  dplyr::left_join(rmcl_predictions, by = c("name", "Region" = "region")) 
```

```{r different seq error}
seq_0.01 <- readRDS(here::here(
  "data-outputs",
  "seq-error",
  "seq_0.01.rds"
)) %>% 
  dplyr::left_join(rmcl_predictions, by = c("name", "Region" = "region"))

seq_0.05 <- readRDS(here::here(
  "data-outputs",
  "seq-error",
  "seq_0.05.rds"
)) %>% 
  dplyr::left_join(rmcl_predictions, by = c("name", "Region" = "region"))

seq_0.1 <- readRDS(here::here(
  "data-outputs",
  "seq-error",
  "seq_0.1.rds"
)) %>% 
  dplyr::left_join(rmcl_predictions, by = c("name", "Region" = "region"))

seq_0.15 <- readRDS(here::here(
  "data-outputs",
  "seq-error",
  "seq_0.15.rds"
)) %>% 
  dplyr::left_join(rmcl_predictions, by = c("name", "Region" = "region"))

seq_0.2 <- readRDS(here::here(
  "data-outputs",
  "seq-error",
  "seq_0.2.rds"
)) %>% 
  dplyr::left_join(rmcl_predictions, by = c("name", "Region" = "region"))
```

```{r bind data }
bind_data <- dplyr::bind_rows(
  list(
    seq_0 = core,
    seq_0.01 = seq_0.01,
    seq_0.05 = seq_0.05,
    seq_0.1 = seq_0.1,
    seq_0.15 = seq_0.15,
    seq_0.2 = seq_0.2
  ), 
  .id = "dataset"
)
```

## Visualizing differences

```{r define some helpful functions, include = FALSE}
plot_bar <- function(data, col, xlab) {
 ggplot(data, aes({{ col }}, fill = dataset)) +
  geom_bar(position = "dodge") +
  scale_fill_viridis_d() +
  labs(x = xlab, y = "Count", fill = "Dataset") +
  theme_coiaf()
}

plot_density <- function(data, col, xlab) {
 ggplot(data, aes({{ col }}, fill = dataset)) +
  geom_density() +
  facet_wrap(~dataset) +
  scale_fill_viridis_d() +
  labs(x = xlab, y = "Density", fill = "Dataset") +
  theme_coiaf()
}

method_comparison <- function(data) {
  pdv <- plot_bar(data, dis_var, "Discrete Variant Method")
  pcv <- plot_density(data, cont_var, "Continuous Variant Method")

  pdf <- plot_bar(data, dis_freq, "Discrete Frequency Method")
  pcf <- plot_density(data, cont_freq, "Continuous Frequency Method")
  
  pdv + pcv + pdf + pcf + plot_layout(guides = "collect")
}

rmcl_helper <- function(data, col) {
  title <- switch(col,
    dis_var = "Discrete Variant Method",
    dis_freq = "Discrete Frequency Method",
    cont_var = "Continuous Variant Method",
    cont_freq = "Continuous Frequency Method"
  )
  
  ggplot(data, aes(x = rmcl_med, y = .data[[col]], color = dataset)) +
    facet_wrap(~dataset) +
    geom_abline(color = "gray", size = 0.5) +
    geom_jitter(alpha = 0.5) +
    theme_bw() +
    theme(
      panel.grid.minor = element_blank(), 
      plot.title = element_text(hjust = 0.5, size = 12), 
      legend.title = element_text(size = 12), 
      legend.position = "right"
    ) +
    guides(size = "none") +
    scale_x_continuous("THE REAL McCOIL Estimation", breaks = seq(0, 10)) +
    labs(title = title, y = "coiaf Estimation", color = "Dataset")
}

rmcl_comparison <- function(data) {
  data = dplyr::filter(data, rmcl_med != 25)
  
  rmcl_helper(data, "dis_var") + 
    rmcl_helper(data, "dis_freq") +
    rmcl_helper(data, "cont_var") +
    rmcl_helper(data, "cont_freq") + 
    plot_layout(guides = "collect")
}
```

```{r data sets methods}
method_comparison(bind_data)
```

```{r data sets rmcl}
rmcl_comparison(bind_data)
```

In order to better visualize the differences between the estimated COI with
differing sequence error, we create one plot examining the discrete Variant
Method.

```{r dvm single plot}
bind_data %>% 
  dplyr::filter(rmcl_med != 25) %>% 
  dplyr::mutate(
    dataset = forcats::as_factor(dataset),
    dataset = forcats::fct_relabel(dataset, function(x) {
      paste("Sequencing Error:", stringr::str_extract_all(x, "(?<=seq_).*"))
    })
  ) %>% 
  ggplot(aes(x = rmcl_med, y = dis_var, color = dataset)) +
  facet_wrap(~dataset) +
  geom_abline(color = "gray", size = 0.5) +
  geom_jitter(alpha = 0.5) +
  theme_coiaf() +
  guides(size = "none", color = "none") +
  scale_x_continuous(
    expression(italic("THE REAL McCOIL")~"Estimation"), 
    breaks = seq(0, 10)
  ) +
  scale_y_continuous(
    expression(italic("coiaf")~"Estimation"), 
    breaks = seq(0, 10)
  )

# Save figure
ggsave(
  here::here("figures", "varying-seq-error.png"),
  device = "png", 
  width = 2000, 
  height = 2000,
  units = "px", 
  dpi = "print"
)
```

## When is the Frequency Method undefined?

Here we aim to understand how to treat samples for which the Frequency Method is
undefined. Our algorithm is undefined when there are too few loci to examine
for the Frequency Method.

First find samples where the Frequency Method is undefined.

```{r high coi samples}
undefined_samples <- core %>%
  dplyr::filter(is.nan(dis_freq), is.nan(cont_freq), rmcl_med == 1)
```

Plot the sample

```{r plot the sample, eval = FALSE}
path <- "~/Desktop/Malaria/COI data/wsafs-coverage/"
plots_per_figure <- 6

purrr::walk(seq_len(3 * 6), function(i) {
  # Assign six plots per figure
  if (i %% plots_per_figure == 1) {
    pdf(paste0("analysis/comparison-figures/coi_3_", i, ".pdf"))
    par(mfrow = c(3, 2))
  }

  # Determine sample details
  # sample_details <- dplyr::slice(undefined_samples, i)
  sample_details <- dplyr::slice_sample(undefined_samples, n = 1)

  # Isolate matrices
  region_matrix <- readRDS(paste0(
    path, 
    glue::glue("wsaf_reg_{ sample_details$Region }.rds")
  ))
  wsaf_matrix <- region_matrix$wsaf
  coverage <- region_matrix$coverage[sample_details$name, ]

  # Determine PLAF and WSAF
  plaf <- colMeans(wsaf_matrix, na.rm = T)
  wsaf <- wsaf_matrix[sample_details$name, ]
  input <- tibble::tibble(wsmaf = wsaf, plmaf = plaf) %>% tidyr::drop_na()

  # Find PLMAF and WSMAF
  minor <- input %>%
    dplyr::mutate(
      wsmaf = ifelse(plmaf > 0.5, 1 - wsmaf, wsmaf),
      plmaf = ifelse(plmaf > 0.5, 1 - plmaf, plmaf)
    ) %>%
    dplyr::filter(wsmaf > 0 & wsmaf < 1)

  # Plot
  plot(
    minor$plmaf, 
    minor$wsmaf, 
    main = sample_details %$% paste0(
      "Sample: ", name,
      "\nV: ", cont_var,
      "; F: ", cont_freq_estimate,
      "; R: ", rmcl_med,
      "; H: ", round(num_variant_loci / expected_num_loci, 2),
      "; C: ", round(mean(coverage, na.rm = TRUE), 2)
    ),
    xlab = "PLMAF",
    ylab = "WSMAF",
    pch = 20, 
    cex = 1, 
    col = "black"
  )
  
  lines(lowess(minor$plmaf, minor$wsmaf), col = "blue", lwd = 2)
  
  # Stop writing to pdf file after 6 plots or after there are no more samples
  if (i %% plots_per_figure == 0 | i == nrow(undefined_samples)) dev.off()
})
```

## Studying the PLMAF

```{r determine PLMAF per region}
path <- "~/Desktop/Malaria/COI data/core-genome/"

region_plmaf <- purrr::map(
  cli::cli_progress_along(seq_len(24), "Finding PLMAF"), 
  function(i) {
    wsaf_matrix <- readRDS(paste0(
      path, 
      glue::glue("wsaf_reg_{ i }.rds")
    ))$wsaf
    
    plaf <- colMeans(wsaf_matrix, na.rm = TRUE)
    ifelse(plaf > 0.5, 1 - plaf, plaf)
})

plot_list <- purrr::map(seq_len(24), function(i) {
  region_plmaf[[i]] %>% 
    tibble::as_tibble() %>% 
    ggplot(aes(x = value)) +
    geom_density() + 
    theme_coiaf()
})

patchwork::wrap_plots(plot_list)
```


## Studying the coverage

```{r determine mean coverage per sample}
path <- "~/Desktop/Malaria/COI data/core-genome/"

sample_coverage <- purrr::map(
  cli::cli_progress_along(seq_len(24), "Finding coverage"), 
  function(i) {
    coverage_matrix <- readRDS(paste0(
      path, 
      glue::glue("wsaf_reg_{ i }.rds")
    ))$coverage
    
    rowMeans(coverage_matrix, na.rm = TRUE)
}) %>% unlist()

mean(sample_coverage, na.rm = TRUE)

sample_coverage %>%
  tibble::as_tibble(rownames = "name") %>%
  dplyr::rename(coverage = value) %>% 
  dplyr::right_join(core) %>% 
  dplyr::group_by(Region) %>% 
  dplyr::summarise(mean_coverage = mean(coverage))
```

```{r determine mean coverage per locus per sample}
coverage <- purrr::map(
  cli::cli_progress_along(seq_len(24), "Finding coverage"), 
  function(i) {
    coverage_matrix <- readRDS(paste0(
      path, 
      glue::glue("wsaf_reg_{ i }.rds")
    ))$coverage
  
    data_region <- dplyr::filter(core, Region == i)
    
    purrr::map_dbl(seq_len(nrow(data_region)), function(j) {
      sample_details <- dplyr::slice(data_region, j)
      sample_coverage <- coverage_matrix[sample_details$name, ]
      round(mean(sample_coverage, na.rm = TRUE), 2)
    })
})

cov_analysis <- core %>%
  tibble::add_column(coverage = unlist(coverage))

cov_analysis %>%
  dplyr::filter(rmcl_med >= 4) %>% 
  ggplot(aes(x = coverage)) +
  geom_histogram(binwidth = 10)

cov_analysis %$%
  cor.test(rmcl_med, coverage)
```

## Studying the number of heterozygous sites

```{r}
het_2pq <- core %>%
  dplyr::filter(is.nan(dis_freq), is.nan(cont_freq), cont_freq_estimate <= 20) %>%
  dplyr::mutate(het_over_2pq = num_variant_loci / expected_num_loci)

# Number variant loci
ggplot(het_2pq, aes(x = het_over_2pq, y = num_variant_loci)) +
  geom_point(alpha = 0.5) +
  stat_density_2d(
    aes(fill = ..level..), 
    geom = "polygon", 
    colour = "white", 
    alpha = 1
  ) +
  facet_wrap(~ rmcl_med) +
  theme_coiaf() +
  guides(fill = "none")

# Cont Variant Method
ggplot(het_2pq, aes(x = het_over_2pq, y = cont_var)) +
  geom_point(alpha = 0.5) +
  stat_density_2d(
    aes(fill = ..level..), 
    geom = "polygon", 
    colour = "white", 
    alpha = 1
  ) +
  facet_wrap(~ rmcl_med, scales = "fixed") +
  theme_coiaf() +
  guides(fill = "none")

# Cont Freq Method
ggplot(het_2pq, aes(x = het_over_2pq, y = cont_freq_estimate)) +
  geom_point(alpha = 0.5) +
  stat_density_2d(
    aes(fill = ..level..), 
    geom = "polygon", 
    colour = "white", 
    alpha = 1
  ) +
  facet_wrap(~ rmcl_med, scales = "fixed") +
  theme_coiaf() +
  guides(fill = "none")

# Cont Variant - Cont Freq
related <- dplyr::filter(het_2pq, rmcl_med == 2)
ggplot(related, aes(x = het_over_2pq, y = cont_freq_estimate - cont_var)) +
  geom_density_2d() +
  geom_point(aes(color = dis_freq_estimate), alpha = 1) +
  facet_wrap(~dis_var, scales = "free_y") +
  theme_coiaf() +
  guides(fill = "none")
```
