---
title: "Pf6 analysis"
author: "Aris Paschalidis"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include = FALSE}
# Here, we set default options for our markdown file
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(coiaf)
library(ggplot2)
library(patchwork)
```

In this analysis file, we examine the result of running our novel algorithms on
the Pf6 dataset. Additionally, we compare the results of our algorithm with the
current state of the art model for predicting COI -- THE REAL McCOIL.

We first load in our data.

```{r load pf6 results}
rmcl_predictions <- readRDS(here::here(
  "data-outputs",
  "rmcl_estimation.rds"
))

complete_predictions <- readRDS(here::here(
  "data-outputs", 
  "core-genome.rds"
)) %>% 
  dplyr::left_join(rmcl_predictions, by = c("name", "Region" = "region"))
```

## Evaluate predictions

We first reshape our data so that we have a single column with our COI
estimates. This involves creating two new additional column to represent the
COI estimation method and optimization method we used to compute the COI.

```{r pivoting data}
pivot_data <- complete_predictions %>%
  tidyr::pivot_longer(
    cols = dplyr::starts_with(c("dis", "cont")),
    names_to = c(".value", "coi_method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::any_of(c("dis", "cont")),
    names_to = "optimization",
    values_to = "COI",
    values_drop_na = TRUE
  ) %>%
  dplyr::relocate(
    dplyr::any_of(c("coi_method", "optimization", "COI")),
    .after = name
  ) %>% 
  dplyr::filter(rmcl_med != 25 & !is.na(rmcl_med))
```

We can next plot all our estimates and further include the deviation from the
median THE REAL McCOIL estimation.

```{r var panel}
var_data <- dplyr::filter(pivot_data, coi_method == "var")

var_panel <- ggplot(
  data = var_data,
  mapping = aes(y = COI, x = rmcl_med, color = optimization)
) +
  scale_size_area() +
  geom_abline(color = "gray", size = 0.5) +
  geom_count(
    data = dplyr::filter(var_data, optimization != "dis"),
    alpha = 0.8,
    position = position_jitter(width = 0.1)
  ) +
  geom_count(
    data = dplyr::filter(var_data, optimization == "dis"),
    alpha = 0.8
  ) +
  theme_coiaf() +
  labs(
    title = "Variant Method",
    y = expression(italic("coiaf")~"Estimation"),
    color = "Optimization Method"
  ) +
  scale_color_manual(
    labels = c("Continuous", "Discrete"),
    values = c("#00239CFF", "#E10600FF")
  ) +
  guides(size = "none") +
    scale_x_continuous(
    expression(italic("THE REAL McCOIL")~"Estimation"), 
    breaks = seq(0, 10)
  )
```

```{r freq panel}
freq_data <- dplyr::filter(pivot_data, coi_method != "var")

freq_panel <- ggplot(
  data = freq_data,
  mapping = aes(y = COI, x = rmcl_med, color = optimization)
) +
  scale_size_area() +
  geom_abline(color = "gray", size = 0.5) +
  geom_count(
    data = dplyr::filter(freq_data, optimization != "dis"),
    alpha = 0.8,
    position = position_jitter(width = 0.1)
  ) +
  geom_count(
    data = dplyr::filter(freq_data, optimization == "dis"),
    alpha = 0.8
  ) +
  theme_coiaf() +
  labs(
    title = "Frequency Method",
    y = expression(italic("coiaf")~"Estimation"),
    color = "Optimization Method"
  ) +
  scale_color_manual(
    labels = c("Continuous", "Discrete"),
    values = c("#00239CFF", "#E10600FF")
  ) +
  guides(size = "none") +
  scale_x_continuous(
    expression(italic("THE REAL McCOIL")~"Estimation"), 
    breaks = seq(0, 10)
  )
```

```{r diff panel}
diff_data <- dplyr::filter(pivot_data, optimization == "cont")

diff_panel <- ggplot(diff_data, aes(x = COI - rmcl_med, fill = coi_method)) +
  geom_density(alpha = 0.6) +
  labs(x = "Difference in Estimation", y = "Density", fill = "COI Method") +
  scale_fill_manual(
    labels = c("Variant", "Frequency"),
    values = c("#4B878BFF", "#D01C1FFF")
  ) +
  theme_coiaf() +
  theme(legend.position = "right")
```

```{r combine panels}
# Use patchwork to combine plots
patchwork <- (var_panel + freq_panel) / diff_panel
patchwork +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r correlation}
wide_data <- pivot_data %>% 
  tidyr::pivot_wider(
    names_from = c(coi_method, optimization),
    names_prefix = "coiaf_",
    values_from = COI
  )

corrr <- wide_data %>% 
  dplyr::select(dplyr::starts_with("coiaf"), rmcl_med) %>% 
  corrr::correlate()

corrr::focus(corrr, rmcl_med)
```

```{r linear regression}
lm(rmcl_med ~ coiaf_var_dis, data = wide_data) %>% broom::glance()
lm(rmcl_med ~ coiaf_var_cont, data = wide_data) %>% broom::glance()
lm(rmcl_med ~ coiaf_freq_dis, data = wide_data) %>% broom::glance()
lm(rmcl_med ~ coiaf_freq_cont, data = wide_data) %>% broom::glance()
```

### COI by region

```{r plot all data}
# Only discrete
ggplot(
  data = dplyr::filter(pivot_data, optimization == "dis"),
  mapping = aes(y = COI, x = rmcl_med, color = coi_method)
) +
  scale_size_area() +
  geom_abline(color = "gray", size = 0.5) +
  geom_count(alpha = 0.7, position = position_jitter(width = 0.1)) +
  facet_wrap(.~ Region, labeller = label_both) +
  scale_x_continuous("THE REAL McCOIL Predictions", breaks = seq(1, 10)) +
  scale_y_continuous("coiaf Predictions", breaks = seq(1, 10)) +
  scale_color_discrete(
    name = NULL, 
    breaks = c("var", "freq"), 
    labels = c("Variant Method", "Frequency Method")
  ) +
  theme_coiaf() +
  guides(size = "none")

# Only continuous
ggplot(
  data = dplyr::filter(pivot_data, optimization == "cont"),
  mapping = aes(y = COI, x = cut_width(rmcl_med, 1), color = coi_method)
) +
  geom_abline(color = "gray", size = 0.5) +
  geom_boxplot() +
  facet_wrap(.~ Region, labeller = label_both) +
  scale_x_discrete("THE REAL McCOIL Predictions", labels = seq(1, 10)) +
  scale_y_continuous("coiaf Predictions", breaks = seq(1, 10)) +
  scale_color_discrete(
    name = NULL, 
    breaks = c("var", "freq"), 
    labels = c("Variant Method", "Frequency Method")
  ) +
  theme_coiaf() +
  guides(size = "none")
```

## World maps

We need to access the Pf6 metadata in order to create a world map as the
metadata contains information about the latitude and longitude of each sample.

```{r load meta}
meta <- readRDS(here::here("metadata", "pf6_meta.rds"))

patient_lat_long <- dplyr::left_join(
  complete_predictions, meta,
  by = c("name" = "Sample")
)
```

```{r sample location summary}
patient_lat_long %>%
  dplyr::count(Continent, Subregion) %>%
  dplyr::mutate(proportion = n / sum(n) * 100) %>%
  dplyr::arrange(Continent, Subregion)
```

We can then generate several world maps with our data!

```{r custom map plot}
# Plot the mean COI
world_map <- function(data,
                      variable,
                      label = NULL,
                      alpha = 0.1,
                      breaks = c(1, 2)) {

  # Access world map data from ggplot2
  world <- ggplot2::map_data("world")

  # Plot world map
  ggplot2::ggplot() +
    ggplot2::borders("world") +
    ggplot2::geom_polygon(
      data = world,
      ggplot2::aes(x = .data$long, y = .data$lat, group = .data$group),
      fill = "grey",
      alpha = 0.5
    ) +
    ggplot2::geom_point(
      data = data,
      ggplot2::aes(
        x = .data$long,
        y = .data$lat,
        size = 1,
        color = {{ variable }}
      ),
      alpha = alpha
    ) +
    ggplot2::scale_color_distiller(
      palette = "YlGnBu", 
      direction = 1,
      limits = c(breaks[1], breaks[length(breaks)]),
      breaks = breaks
    ) +
    ggplot2::theme_void() +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::scale_size(guide = "none") +
    ggplot2::labs(color = label) +
    ggplot2::coord_quickmap(xlim = c(-75, 150), ylim = c(-30, 30))
}
```


```{r world map, warning = FALSE, fig.height = 2}
# Get average of data
map_average <- patient_lat_long %>%
  dplyr::group_by(Region, Lat, Long) %>%
  dplyr::summarise(
    coi_mean = mean(dis_var, na.rm = TRUE),
    coi_med = median(dis_var, na.rm = TRUE),
    prev = median(as.numeric(prev_2_10)),
    rmcl_med = median(rmcl_med),
    .groups = "drop"
  ) %>%
  dplyr::rename(lat = Lat) %>%
  dplyr::rename(long = Long)

# Plot the mean COI
mean_world <- world_map(
  map_average,
  coi_mean,
  label = "Mean COI",
  alpha = 0.7,
  breaks = c(1.0, 1.5, 2.0, 2.5)
)
mean_world

# Plot the median COI
med_world <- world_map(
  map_average,
  coi_med,
  label = "Median COI",
  alpha = 0.7,
  breaks = c(1.0, 1.5, 2.0)
)
med_world
```

```{r all patients map, warning = FALSE, fig.height = 2}
# Get data that has a COI of less than 10. This is done for now as we suspect
# there are some samples that are wrong! Regions that predict 25!
# The two patients are: "PN0075-C" and "PT0069-C"
# We also set up data to be descending so that smaller COIs stay on top.
all_world_map <- patient_lat_long %>%
  dplyr::arrange(desc(dis_var)) %>%
  dplyr::rename(lat = Lat) %>%
  dplyr::rename(long = Long)

# Plot all patient COIs (theme is added so that there is less whitespace)
all_world <- world_map(
  all_world_map,
  dis_var,
  "COI",
  alpha = 0.7,
  breaks = c(1, 2, 4, 6, 8, 10)
)
all_world
```

```{r kruskal-wallis}
# Mean, median COI per continent
patient_lat_long %>%
  dplyr::group_by(Continent) %>% 
  dplyr::summarise(
    mean = mean(dis_var, na.rm = TRUE), 
    median = median(dis_var, na.rm = TRUE),
    .groups = "drop"
  )

# KW across 24 regions
kruskal.test(dis_var ~ Continent, data = patient_lat_long)

# KW across subregions of continents
kruskal.test(dis_var ~ Subregion, data = patient_lat_long)
```

```{r nemenyi continent}
# Post hoc Nemenyi test
pvalue_continent <- PMCMRplus::kwAllPairsNemenyiTest(
  dis_var ~ as.factor(Continent), 
  data = patient_lat_long
)$p.value

purrr::map_df(tibble::as_tibble(pvalue_continent), ~ round(.x, digits = 5)) %>% 
  tibble::add_column(Continent = rownames(pvalue_continent), .before = 1)
```

```{r subregion}
patient_lat_long %>%
  dplyr::group_by(Continent, Subregion) %>% 
  dplyr::summarise(
    mean = mean(dis_var, na.rm = TRUE),
    sd = sd(dis_var, na.rm = TRUE),
    median = median(dis_var, na.rm = TRUE),
    iqr = IQR(dis_var, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

pvalue_subregion <- PMCMRplus::kwAllPairsNemenyiTest(
  dis_var ~ as.factor(Subregion), 
  data = patient_lat_long
)$p.value

purrr::map_df(tibble::as_tibble(pvalue_subregion), ~ round(.x, digits = 5)) %>% 
  tibble::add_column(Subregion = rownames(pvalue_subregion), .before = 1)
```


## Prevalence & FwS

We can also look at the relationship between our COI estimations and the FwS
metric as well as the prevalence of mutations.

```{r prev fws data}
prev_fws <- patient_lat_long %>%
  dplyr::filter(rmcl_med != 25 & !is.na(rmcl_med)) %>%
  tidyr::pivot_longer(
    cols = c("dis_var", "cont_var", "dis_freq", "cont_freq", "rmcl_med"),
    names_to = "Method",
    values_to = "COI"
  ) %>%
  dplyr::relocate(dplyr::any_of(c("Method", "COI")), .after = name) %>% 
  dplyr::mutate(
    Method = dplyr::recode(
      Method,
      dis_var = "Discrete \n Variant Method",
      cont_var = "Continuous \n Variant Method",
      dis_freq = "Discrete \n Frequency Method",
      cont_freq = "Continuous \n Frequency Method",
      rmcl_med = "THE REAL McCOIL"
    )
  )
```

### FwS

```{r fws}
ggplot(data = prev_fws, aes(x = Fws, y = COI, color = Method)) +
  geom_point(alpha = 0.3, position = position_jitter(height = 0.2)) +
  labs(x = "FwS", y = "Estimated COI") +
  facet_grid(~Method) +
  guides(color = "none") +
  theme_coiaf()

# Correlation
coiaf_fws <- dplyr::filter(prev_fws, Method == "Discrete \n Variant Method")
cor.test(coiaf_fws$Fws, coiaf_fws$COI)

rmcl_fws <- dplyr::filter(prev_fws, Method == "THE REAL McCOIL")
cor.test(rmcl_fws$Fws, rmcl_fws$COI)
```

### Prevalence

```{r prev, warning = FALSE}
ggplot(data = prev_fws, aes(x = prev_2_10, y = COI, color = Method)) +
  geom_point(alpha = 0.2, position = position_jitter(height = 0.2)) +
  geom_smooth(method = "lm", formula = y ~ x) +
  scale_x_log10(
    breaks = c(0.01, 0.1, 1),
    labels = c(0.01, 0.1, 1),
    limits = c(0.01, 1)
  ) +
  facet_grid(~forcats::as_factor(Method)) +
  labs(x = "Log10 Prevalence", y = "Estimated COI") +
  guides(color = "none") +
  theme_coiaf()

ggplot(
  data = prev_fws, 
  aes(
    x = cut(prev_2_10, seq(0, 1, 0.1), right = TRUE),
    y = COI, color = Method
  )
) +
  geom_boxplot(alpha = 0.3) +
  facet_grid(~forcats::as_factor(Method)) +
  labs(x = "Prevalence", y = "Estimated COI") +
  guides(color = "none") +
  theme_coiaf()
```

We could also look at the prevalence of mutations slightly differently and
look at the ridge plot. Here, we no longer compare our data to THE REAL McCOIL
so we do not want to remove any observations from our data.

```{r prevalence}
global_prev <- patient_lat_long %>%
  dplyr::select(!dplyr::starts_with("rmcl")) %>%
  tidyr::pivot_longer(
    cols = c("dis_var", "cont_var", "dis_freq", "cont_freq"),
    names_to = c(".value", "coi_method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::any_of(c("dis", "cont")),
    names_to = "optimization",
    values_to = "COI",
    values_drop_na = TRUE
  ) %>%
  dplyr::relocate(
    dplyr::any_of(c("coi_method", "optimization", "COI")),
    .after = name
  )

continents <- global_prev %>%
  dplyr::filter(coi_method == "var" & optimization == "dis") %>%
  dplyr::group_by(Region) %>%
  dplyr::mutate(med_prev = median(prev_2_10, na.rm = TRUE)) %>%
  dplyr::filter(med_prev != 0) %>%
  dplyr::arrange(prev_2_10) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    Region = forcats::as_factor(Region),
    Subregion = forcats::as_factor(Subregion),
    Subregion = forcats::fct_relevel(
      Subregion, "Western Africa", "Middle Africa", "Eastern Africa", 
      "South America", "Melanesia", "South-Eastern Asia", "Southern Asia"
    )
  )

sorted_prev <- continents %>%
  dplyr::select(Region, med_prev) %>%
  dplyr::distinct() %>%
  dplyr::pull(med_prev) %>%
  sort()

ridges <- ggplot(
  data = continents,
  aes(y = forcats::fct_reorder(Region, med_prev), x = COI, fill = Subregion)
) +
  ggridges::geom_density_ridges() +
  ggridges::theme_ridges() +
  theme_coiaf() +
  scale_fill_manual(
    values = c("#31a354", "#a1d99b", "#e5f5e0", "#756bb1", "#fdbe85", "#9ecae1", "#deebf7")
  ) +
  annotate(
    "text",
    x = 7.35,
    y = seq_len(length(sorted_prev)) + 0.4,
    label = round(sorted_prev, 3),
    size = 2.2
  ) +
  labs(x = "COI", y = "Region")
```

Combining our world map plots and our prevalence plot, we can generate an
informative picture of COI in the world.

```{r coi around the world, message = FALSE, warning = FALSE}
mean_world / med_world / ridges +
  plot_annotation(tag_levels = "A") +
  plot_layout(widths = c(2, 2, 1), heights = c(1, 1, 1.5))
```

We also can generate some stats to examine the relationship between prevalence
and the COI.

```{r prev in each region}
patient_lat_long %>%
  dplyr::group_by(Continent) %>% 
  dplyr::summarise(
    median = median(prev_2_10, na.rm = TRUE),
    coi = mean(dis_var, na.rm = TRUE),
    .groups = "drop"
  )

pvalue_subregion <- PMCMRplus::kwAllPairsNemenyiTest(
  prev_2_10 ~ as.factor(Continent), 
  data = patient_lat_long
)$p.value

purrr::map_df(tibble::as_tibble(pvalue_subregion), ~ round(.x, digits = 5)) %>% 
  tibble::add_column(Subregion = rownames(pvalue_subregion), .before = 1)
```

```{r prev correlation}
patient_lat_long %>%
  dplyr::select(dis_var, cont_var, dis_freq, cont_freq, rmcl_med, prev_2_10) %>%
  corrr::correlate() %>% 
  corrr::focus(prev_2_10)
```

```{r prev lm}
regressions <- global_prev %>%
  dplyr::select(COI, prev_2_10, coi_method, optimization) %>% 
  tidyr::nest(data = c(COI, prev_2_10)) %>% 
  dplyr::mutate(
    fit = purrr::map(data, ~ lm(COI ~ prev_2_10, data = .x)),
    tidied = purrr::map(fit, broom::tidy),
    glanced = purrr::map(fit, broom::glance),
    augmented = purrr::map(fit, broom::augment)
  )

tidyr::unnest(regressions, tidied)
tidyr::unnest(regressions, glanced)
```

```{r coi with varying prev thresholds}
prev_leq <- patient_lat_long %>% 
  dplyr::filter(prev_2_10 <= 0.01) %>% 
  dplyr::pull(dis_var) %>% 
  table() %>% 
  prop.table() %>% 
  round(3)
sum(prev_leq[1:2]) * 100

prev_geq <- patient_lat_long %>% 
  dplyr::filter(prev_2_10 >= 0.1) %>% 
  dplyr::pull(dis_var) %>% 
  table() %>% 
  prop.table() %>% 
  round(3)
(1 - sum(prev_geq[1:2])) * 100
```
